\noindent\rule{13cm}{1.0pt}\\ \vspace{0.2cm} A Polynomial-time Reduction of the 3-SAT to the Quadratic Congruence and other Related Problems \noindent\rule{13cm}{0.8pt}

### 2017-11-19 paper/paper.tex
Second draft.

German mathematician David Hilbert published~\cite{hilbert} in 1902 a set of 23 unsolved problems
in mathematics he deemed to be the most important mathematical problems of the 20th century.  Since
then 9 of them have been solved (at the time the author is writing this line and as far as the
author is aware), 9 are considered partially solved, three of them are unsolved and two of them are
considered too vague. Unsolved problems include the infamous Riemann Hypothesis and an extension to
the Kronecker-Weber Theorem. Amongst solved problems is the 10th Hilbert problem.

@@

It was answered in 1970 by Matiyasevich~\cite{diophantine} to be impossible. The question now
becomes, in which cases is there an algorithm for solvability and what is the complexity of such
algorithms? In 1976, Adleman and Manders~\cite{qcp1} partially answered these questions by proving
that, for the quadratic cases, there exists an algorithm and it is NP-complete. In their proof,
they also found that, through a slight modification in the final step of their proof, it was
possible to answer the quadratic congruence problem. A cleaner version of this proof was published
in 1978 by Adleman and Manders~\cite{qcp2}, proof we try to explain in this paper.

In this paper we focus on the second result of Adleman and Manders' 1978 article, but also show the
main result, namely that the set of quadratic diophantine equations with natural numbers solutions
is NP-complete. The proof is done through a polynomial-time reduction from the 3-SAT problem. This
reduction implies that both problems covered in Adleman and Manders' article are in NP-complete.

In 1971, American mathematician Stephen Cook published ``The Complexity of Theorem-proving
Procedures''~\cite{cook}, and in the next year, his fellow countryman Richard Karp published
``Reducibility Among Combinatorial Problems''~\cite{karp}. The two articles introduced the concepts
of P and NP classes, yielding the duo a Turing Award. Interestingly in 1973, on the other side of
the Iron Curtain, Ukrainian Leonard Levin published~\cite{levin} in the USSR equivalent results to
Cook's and Karp's, but considering search problems instead of decision problems (an interesting
remark is that Levin did not receive a Turing Award for his work, despite having achieved
equivalent results). Both works resulted in the following statement: that any problem in NP can be
reduced in polynomial time by a deterministic Turing machine to the problem of satisfiability of a
Boolean formula, i.e.\ the SAT problem.  Additionally, if there exists a deterministic polynomial
time algorithm for solving SAT, then every NP problem can be solved by a deterministic polynomial
time algorithm.

@@

@@

In his 1972 paper, Karp also published a list of 21 NP-complete problems in which he showed
reductions implying its membership in the NP-complete class. The list above is Karp's 21
NP-complete problems, where the nesting indicates the direction of the reduction. For instance, the
exact cover problem was reduced to the knapsack problem, chromatic number was reduced to exact
cover, 3-SAT was reduced to exact cover, and the SAT was reduced to the 3-SAT problem.

From this we know that the 3-SAT is NP-complete. Not only that, any problem that reduces 3-SAT is
also in NP\@. In the next sections we will provide a brief review on polynomial-time reductions,
present proper definitions on the QCP and 3-SAT, and prove the reduction. This section provided a
brief historical overview, but we do not restrict historical remarks to only this section. When
relevant, we provide short historical anecdotes on the subject.

@@

It is natural to conclude that problems have a certain complexity attached to them. Furthermore, we
can provide several algorithms that take the same input and yield the same output. In this case in
particular we know that sorting a vector is $\bigo(n\ln n)$ at best. Examples of algorithms for the
above problem are InsertSort, MergeSort and BubbleSort, with each one having different worst case
complexities. We next show a few examples of different types of problems:

@@

Let $\Pi$ be a Problem. We say that $\Pi$ admits a polynomial verifier $\verifier$ for a YES answer
if there exists a polynomial algorithm that takes an instance $I$ of $\Pi$ and an object
$\mathcal{C}$ such that the size of $\mathcal{C}$ is polynomial and returns YES for some
$\mathcal{C}$ if the answer $\Pi(I)$ is YES and NO for all $\mathcal{C}$ if $\Pi(I)$ is NO\@.

@@

  \begin{align*}
    &(x_p\vee x_q\vee x_l): &y_k=r(x_p)+r(x_q)+r(x_l)-1\implies &-1\leq y_k\leq 2\\
    &(\ov{x}_p\vee x_q\vee x_l): &y_k=r(x_q)+r(x_l)+1-r(x_p)-1\implies &-1\leq y_k\leq 2\\
    &(\ov{x}_p\vee\ov{x}_q\vee x_l):&y_k=r(x_l)+2-r(x_p)-r(x_q)-1\implies &-1\leq y_k\leq 2\\
    &(\ov{x}_p\vee\ov{x}_q\vee\ov{x}_l):&y_k=3-r(x_p)-r(x_q)-r(x_l)-1\implies &-1\leq y_k\leq 2
  \end{align*}

  Note how in all cases above, the clause is satisfiable if and only if $y_k\neq -1$. In the first
  case, when $y_k=-1$, no literal $x_i$ is valuated as true. In the second case, when $y_k\neq -1$,
  no literal $x_i$ is valuated as true and $\ov{x}_p$ is valuated as false. The third case is
  similar, with $x_l$ false, and $\ov{x}_p,\ov{x}_q$ false. For the last case, all $\ov{x}_i$ are
  set to false. When $y_k\neq -1$, there exists some literal being valuated as true, and therefore
  the clause $\sigma_k$ is valuated as true.

@@

  \begin{align*}
    &(x_p\vee x_q\vee x_l): &y_k=r(x_p)+r(x_q)+r(x_l)\implies &0\leq y_k\leq 3\\
    &(\ov{x}_p\vee x_q\vee x_l): &y_k=r(x_q)+r(x_l)+1-r(x_p)\implies &0\leq y_k\leq 3\\
    &(\ov{x}_p\vee\ov{x}_q\vee x_l):&y_k=r(x_l)+2-r(x_p)-r(x_q)\implies &0\leq y_k\leq 3\\
    &(\ov{x}_p\vee\ov{x}_q\vee\ov{x}_l):&y_k=3-r(x_p)-r(x_q)-r(x_l)\implies &0\leq y_k\leq 3
  \end{align*}

>>>>

German mathematician David Hilbert published in 1902~\cite{hilbert} a set of 23 unsolved problems
in mathematics he deemed to be the most important mathematical problems to be solved in the 20th
century. Since then 9 of them have been solved, 9 are considered partially resolved, three of them
are unsolved and two of them are considered too vague (as of the time the author is writing this
line and as far as the author is aware). Unsolved problems include the infamous Riemann Hypothesis
and an extension to the Kronecker-Weber Theorem. Amongst solved problems is the 10th Hilbert
problem.

@@

It was answered in 1970 by Matiyasevich to be impossible~\cite{diophantine}. The question now
becomes, in which cases is there an algorithm for solvability and what is the complexity of such
algorithm? In 1976, Adleman and Manders~\cite{qcp1} partially answered these questions by proving
that, for the quadratic case, there exists an algorithm and the problem of finding the solutions is
NP-complete. In their proof, they also found that, through a slight modification in the final step
of their proof, it was possible to answer the quadratic congruence problem. A cleaner version of
this proof was published in 1978 by Adleman and Manders~\cite{qcp2}, proof we try to explain in
this paper.

In this paper we focus on the second result of Adleman and Manders' 1978 article (namely that the
quadratic congruence problem is NP-complete), but also briefly show the main result, i.e\ that
the set of quadratic diophantine equations with natural numbers solutions is NP-complete. The proof
is done through a polynomial-time reduction from the 3-SAT problem. This reduction implies that
both problems covered in Adleman and Manders' article are NP-complete.

In 1971, American mathematician Stephen Cook published ``The Complexity of Theorem-proving
Procedures''~\cite{cook}, and in the next year, his fellow countryman Richard Karp published
``Reducibility Among Combinatorial Problems''~\cite{karp}. The two articles introduced the concepts
of P and NP classes, yielding the duo a Turing Award. Interestingly in 1973, on the other side of
the Iron Curtain, Ukrainian Leonard Levin published~\cite{levin} in the USSR equivalent results to
Cook's and Karp's, but considering search problems instead of decision problems (an interesting
remark is that Levin did not receive a Turing Award for his work, despite achieving equivalent
results). Both works resulted in the following statement: that any problem in NP can be reduced in
polynomial time, in a deterministic Turing machine, to the problem of satisfiability of a Boolean
formula, i.e.\ the SAT problem.  Additionally, if there exists a deterministic polynomial time
algorithm for solving SAT, then every NP problem can be solved by a deterministic polynomial time
algorithm.

@@

In his 1972 paper, Karp also published a list of 21 NP-complete problems, showing the
polynomial-time reductions of 21 problems. Below is Karp's list, where the nesting indicates the
direction of the reduction. For instance, the exact cover problem was reduced to the knapsack
problem, chromatic number was reduced to exact cover, 3-SAT was reduced to exact cover, and the SAT
was reduced to the 3-SAT problem.

@@

From this we know that one can reduce 3-SAT to any problem that the latter will be proved to be
NP-complete. In the next sections we will provide a brief review on polynomial-time reductions,
present proper definitions on the QCP and 3-SAT, and finally prove the reduction. The last section
is devoted to related problems. We show the Diophantine problem presented in~\cite{qcp2} and other
Number Theory problems reductions.

@@

It is natural to conclude that problems have a certain complexity attached to them. Furthermore, we
can provide several algorithms that give answers to our problems. In this case in particular we
know that sorting a vector is $\bigo(n\lg n)$ at best. Examples of algorithms for the above problem
are InsertSort, MergeSort and BubbleSort, with each one having different worst case complexities.
We next show a few examples of different types of problems:

@@

Let $\Pi$ be a Problem. We say that $\Pi$ admits a polynomial verifier $\verifier$ for a YES answer
if there exists a polynomial algorithm that takes an instance $I$ of $\Pi$ and an object
$\mathcal{C}$ such that the size of $\mathcal{C}$ is polynomial in $I$ and returns YES for some
$\mathcal{C}$ if the answer $\Pi(I)$ is YES and NO for all $\mathcal{C}$ if $\Pi(I)$ is NO\@.

@@

  \begin{align*}
    &(x_p\vee x_q\vee x_t): &y_k=r(x_p)+r(x_q)+r(x_t)-1\implies &-1\leq y_k\leq 2\\
    &(\ov{x}_p\vee x_q\vee x_t): &y_k=r(x_q)+r(x_t)+1-r(x_p)-1\implies &-1\leq y_k\leq 2\\
    &(\ov{x}_p\vee\ov{x}_q\vee x_t):&y_k=r(x_t)+2-r(x_p)-r(x_q)-1\implies &-1\leq y_k\leq 2\\
    &(\ov{x}_p\vee\ov{x}_q\vee\ov{x}_t):&y_k=3-r(x_p)-r(x_q)-r(x_t)-1\implies &-1\leq y_k\leq 2
  \end{align*}

  Note how in all cases above, the clause is satisfiable if and only if $y_k\neq -1$. In the first
  case, when $y_k=-1$, no literal $x_i$ is valuated as true. In the second case, when $y_k\neq -1$,
  no literal $x_i$ is valuated as true and $\ov{x}_p$ is valuated as false. The third case is
  similar, with $x_t$ false, and $\ov{x}_p,\ov{x}_q$ false. For the last case, all $\ov{x}_i$ are
  set to false. When $y_k\neq -1$, there exists some literal being valuated as true, and therefore
  the clause $\sigma_k$ is valuated as true.

@@

  \begin{align*}
    &(x_p\vee x_q\vee x_t): &y_k=r(x_p)+r(x_q)+r(x_t)\implies &0\leq y_k\leq 3\\
    &(\ov{x}_p\vee x_q\vee x_t): &y_k=r(x_q)+r(x_t)+1-r(x_p)\implies &0\leq y_k\leq 3\\
    &(\ov{x}_p\vee\ov{x}_q\vee x_t):&y_k=r(x_t)+2-r(x_p)-r(x_q)\implies &0\leq y_k\leq 3\\
    &(\ov{x}_p\vee\ov{x}_q\vee\ov{x}_t):&y_k=3-r(x_p)-r(x_q)-r(x_t)\implies &0\leq y_k\leq 3
  \end{align*}
